Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-07 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
Section II.C

The QNMs of the Schwarzschild BH in dCS gravity were studied in
Refs.~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb}, that found that scalar
perturbations and gravitational perturbations of axial parity couple. (Contrast
with EdGB gravity).
%
This yields in a breakdown of isospectrality. The QNM spectra of slowly-rotating BHs in dCS gravity was
studied in Ref.~\cite{Wagle:2021tam}.
%
They were also extracted from numerical relativity (NR) simulations of head-on BH collisions
in Ref.~\cite{Okounkova:2019dfo}.

@@

Our last example of modified gravity theories are the so-called effective field
theories (EFT) of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell$ is a lengthscale, assumed to be small compares to the lengthscale
associated with a BH, i.e., $\ell / M \ll 1$, and $L^{(2n)}$ are corrections to
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up to dimension-eight operators
($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and where $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ ($i=1,2,3$) are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we will focus on a subset of the theory's
parameter space.
%
In particular, we will consider dimension-six and dimension-eight operators separately.
%
In the dimension-six case we further assume that $\lame = \lamo = 1$, leaving us with $\ell = \ell_{\rm cEFT}$
as our single free parameter.
%
In the dimension-eight case, we will set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell = \ell_{\rm qEFT}$ as our single free parameter.

@@

The QNMs of nonrotating BHs where calculated in in Refs.~\cite{deRham:2020ejn,Cano:2020cao} in the dimension-six EFT
and in Refs.~\cite{Cardoso:2018ptl,Cano:2020cao} in the dimension-eight EFT.
%
Ref.~\cite{Cano:2020cao} calculated the leading-order in BH spin corrections to the QNM spectra.

>>>>

The QNMs of the Schwarzschild BH in dCS gravity were studied in
Refs.~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb}, that found that scalar
perturbations couple to gravitational perturbations of axial parity. (Contrast
with EdGB gravity)\agcomm{Is the text in parenthesis a comment?}.
%
This leads to a breakdown of isospectrality. The QNM spectra of slowly-rotating BHs in dCS gravity was
studied in Ref.~\cite{Wagle:2021tam}.
%
They were also extracted from numerical relativity (NR) simulations of head-on BH collisions
in Ref.~\cite{Okounkova:2019dfo}.

@@

Our last example of modified gravity theories are the so-called effective field
theories (EFT) of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell$ is a length-scale assumed to be small compared to the length-scale
$\ell / M \ll 1$ associated with a BH and $L^{(2n)}$ are corrections 
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up to dimension-eight operators
($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and $(\lambda_{\rm o, e},\varepsilon_{i})$,  $i=1,2,3$, are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we will focus on a subset of the parameter space.
%
In particular, we will consider dimension-six and dimension-eight operators separately.
%
In the dimension-six case we further assume that $\lame = \lamo = 1$, leaving us with $\ell = \ell_{\rm cEFT}$
as our single free parameter.
%
In the dimension-eight case, we will set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell = \ell_{\rm qEFT}$ as our single free parameter \agcomm{Perhaps a bit of motivation about why we make these choices, even if it is to say that it is a simple first choice?}.

@@

The QNMs of nonrotating BHs where calculated in Refs.~\cite{deRham:2020ejn,Cano:2020cao} in the dimension-six EFT
and in Refs.~\cite{Cardoso:2018ptl,Cano:2020cao} in the dimension-eight EFT.
%
Ref.~\cite{Cano:2020cao} calculated the leading-order BH spin corrections to the QNM spectra.
